rapr: Representative and Adequate Prioritisations in R

### 2016-06-15 code/rmarkdown/article.Rmd
+ fixed bug in makefile + finished figures + made progress in methods

## Methods
To investigate the utility and beaviour of this problem, we conducted a simulation study and two case-studies. These studies involved generating prioritizations using only amount-based targets to represent prioritisations generated using conventional methods (eg. Marxan), and prioritizations using amount-based and space-based targets. All prioritisations were solved to within `r (general.params.LST[[MODE]]$MIPGap * 100)` % of optimality.

### Simulation study
We generated a study area with planning units arranged in $`r sqrt(sim.params.LST[[MODE]]$number.planning.units)` \times `r sqrt(sim.params.LST[[MODE]]$number.planning.units)`$ square grid. We then simulated three species. The first species was simulated to represent a hyper-generalist--occurring in all planning units with equal probability. This species' distribution was based on a uniform distribution (eqn 8a). The second species was simulated to represent a species with an idealised distribution. It has a core range area, and is less likely to be found in areas that are distant from the core area. This species' distribution was simulated using the probability density function of a single multivariate normal distribution (eqn 8b). The third species represents a species with two distinct populations. This species' distribution was simulated to depict a bimodal distribution, based on the combination of the probability density functions of two multivariate normal distributions (eqn 8c). The probability that each species inhabited a given planning unit was calculated using the $X, Y$ coordinates of the planning unit and eqns 8a--8c. We used a geographic attribute space to showcase the behavior of the problem. For each species, demand points were generated by calculating the centroids of each planning units and their weights were set as the probability that the planning units were occupied.

\begin{align*}
& P (\text{uniform species} | x, y}) &= 0.1 & \tag{8a}\\
& P (\text{normal species} | x, y}) &= XXXX & \tag{8b}\\
& P (\text{uniform species} | x, y}) &= XXX & \tag{8c}\\
\end{align*}

We generated four prioritisations for each species. First to represent prioritisations generated using conventional conservation planning methods, we generated prioritisations using only `` % amount targets. Second to show how the addition of geographic targets can affect a prioritisation, we  generated prioritisations using `r (sim.params.LST[[MODE]]$amount.target*100)` % amount targets and `r (sim.params.LST[[MODE]]$space.target*100)` % geographic targets. Third to represent prioritisations using conventional planning methods that penalise for fragmentation, we generated prioritisations using `r (sim.params.LST[[MODE]]$amount.target*100)` % amount targets and a boundary length modifier of `r sim.params.LST[[MODE]]$blm`. Fourth to illustrate the combined affect using geographic targets and penalising fragmentation, we generated prioritisations using using `r (sim.params.LST[[MODE]]$amount.target*100)` % amount targets and `r (sim.params.LST[[MODE]]$space.target*100)` % geographic targets and a boundary length modifier of `r sim.params.LST[[MODE]]$blm`.

### Case-study 1
Here we investigated the potential benefits of using space-based targets to secure a representative proportion of a species' realised niche. We selected Queensland, Australia as the study area. This region is ideal for exploring the potential of niche-based targets because it contains a broad range of different habitats--such as rainforests, woodlands, and deserts. We obtained data for 19 bioclimatic variables across the region (at XXX resolution from XXXX; XXXX) and subject them to a principle components analysis to obtain the main gradients of environmental variation. We used the first two principle components (cumulatively explained `r ` % of the total variation) to characterise the environental variation across Queensland, Australia.

We used five bird species in this case study: blue-winged kookaburra (XXX), brown-backed honeyeater (XXX), brown falcon (XXX), and the southern cassory (XXXX)  . These bird species have different range-sizes and life-history strategies. For instance XXXXX is a generalist and occurs across most of Australia, whereas XXXX is a specialist and is only found in the rainforests of north-east Queensland. For each species, we mapped their distriubtion using the following method: we obtained observation data from the Atlas of Living Australia across the whole of Australia (using the \texttt{ALA4R R} package), spatially thinned the data to omit points within $10 km^2$ of each other to ameloriate the effects of sampling bias (using the developmental version of \texttt{spThin R} package; XXXX), and fit 85 % minimum convex polygons (using the \texttt{adehabitatHR R} package; XXX). We used this method because it is reproducible and moderately accurate. We generated `r ` demand points for each species using hyperbox kernels (implemented in the \texttt{hypervolume R} package; XXXX).

We generated two prioritisations. The first prioritisation was generated using `r ` % amount-based targets for all species. The second prioritisation was generated the same amount-based target with an additional ` r ` % niche-based for each species.

### Case-study 2

## Results
### Simulated species

### Case-study 1

### Case-study 2

## Implications and future directions
The \texttt{rapr} R package provides a unified approach to reserve selection. This R package provides decision makers with the tools to generate prioritisations that secure both biodiversity patterns and processes. Additionally, the package contains funcionality to accommodate uncertainty in the distribution of features, and also identify suitably connected reserves. Both the simulated and case-study species suggest that conservation planning exercises need to explicitly consider biodiversity processes during the reserve selection process to capture them.

One of the key advantages of the \texttt{rapr R} package is that it is general enough that any spatial variation could be considered an attribute space, regardless of whether this variation is intrinsic or extrinsic to the feature(s). For example, advances in genomic fields produced high resolution data on genetic information (eg. amplified fragment length polymorphisms, AFLPs; single nucleotide polymorphisms, SNPs). By using geostatistical analysis [eg. generalised dissimilarity modelling GDMs and gradient forests; @r370; @r452], this data has been used to generate maps describing the spatial distribution of genomic variation within a species [@r361; @r419]. These maps in turn could be used to construct a genomic attribute space, and in turn, could be used to generate prioritisations that secure a representative sample of genomic variation within a species. However, because the problem formulations used in this package are so general, the tools in this package could be misused, and generate poor quality prioritisations.

@@

The \texttt{rapr R} package could be further extended to identify more effective prioritisations. First, the formulation of fragmentation used in this package may be too simplistic in some cases (eg. exercises involving multiple species with different dispersal capabilities), and more realistic measures of fragmentation (eg. those used in Zonation) could be used to identify more effective prioritisations. Second, the problem does not consider temporal dynamics. Here, conservation actions are assumed to be implemented simultaneously in all selected planning units and assumed to remain implemented for all time. As a consequence, this \texttt{rapr R} package is not useful for scenarios where actions are implemented during multiple discrete periods in time (eg. actions are made adue to annual funding cycles), or scenarios involving threatening processes that vary across space and time [reviewed in @r11]. Future research may look into incorporating such elements into this \texttt{R} package.

To maximise the long-term persistence of biodiversity--the stated goal of conservation--decision makers need to identify prioritisations that preserve existing patterns of biodiversity and the processes that support them. To achieve this, conservation planners need a decision support tool that can explicitly accommodate biodiversity patterns and processes. Here, we developed the \texttt{rapr R} package to fill this void. By exploring the functiontality of this package using several simulated species, we found that including space-based targets can radically change a prioritisation for the simplest of species.

>>>>

To understand the beaviour of this problem and showcase its utility, we conducted a simulation study and two case-studies. These studies involved generating prioritizations using only amount-based targets to represent prioritisations generated using conventional methods (eg. Marxan), and prioritizations using amount-based and space-based targets using the unreliable formulation.

## Simulation study
### Methods
We generated a study area with planning units arranged in $`r sqrt(sim.params.LST[[MODE]]$number.planning.units)` \times `r sqrt(sim.params.LST[[MODE]]$number.planning.units)`$ square grid (Figure 2). We then simulated three species. The first species was simulated to represent a hyper-generalist--occurring in all planning units with equal probability (Figure 2a). This species' distribution was based on a uniform distribution (eqn 8a). The second species was simulated to represent a species with an idealised distribution (Figure 2b). It has a core range area, and is less likely to be found in areas that are distant from the core area. This species' distribution was simulated using the probability density function of a single multivariate normal distribution (eqn 8b). The third species represents a species with two distinct populations (Figure 2c). This species' distribution was simulated to depict a bimodal distribution, based on the combination of the probability density functions of two multivariate normal distributions (eqn 8c). The probability that each species inhabited a given planning unit was calculated using the $X, Y$ coordinates of the planning unit and eqns 8a--8c. We used a geographic attribute space to showcase the behavior of the problem. For each species, demand points were generated by calculating the centroids of each planning units and their weights were set as the probability that the planning units were occupied.

\begin{align*}
& P (\text{uniform species} | x, y) &= 0.1 & \tag{8a}\\
& P (\text{normal species} | x, y) &= XXXX & \tag{8b}\\
& P (\text{uniform species} | x, y) &= XXX & \tag{8c}\\
\end{align*}

We generated four prioritisations for each species. First to represent prioritisations generated using conventional conservation planning methods, we generated prioritisations using only `r sim.params.LST[[MODE]]$amount.target*100` % amount targets. Second to show how the addition of geographic targets can affect a prioritisation, we  generated prioritisations using `r (sim.params.LST[[MODE]]$amount.target*100)` % amount targets and `r (sim.params.LST[[MODE]]$space.target*100)` % geographic targets. Third to represent prioritisations using conventional planning methods that penalise for fragmentation, we generated prioritisations using `r (sim.params.LST[[MODE]]$amount.target*100)` % amount targets and a boundary length modifier of `r sim.params.LST[[MODE]]$blm`. Fourth to illustrate the combined affect using geographic targets and penalising fragmentation, we generated prioritisations using using `r (sim.params.LST[[MODE]]$amount.target*100)` % amount targets and `r (sim.params.LST[[MODE]]$space.target*100)` % geographic targets and a boundary length modifier of `r sim.params.LST[[MODE]]$blm`.

### Results

## Case-study 1
### Methods
Here we investigated how space-based targets can be used in a multi-species planning context to generate a prioritisation that sufficiently preserves the realised niche for several species. We selected Queensland, Australia as the study area. This region is ideal for exploring the potential of niche-based targets because it contains a broad range of different habitats. We obtained data for 19 bioclimatic variables across the region [at $30^{\prime \prime}$ resolution from \url{www.worldclim.org}; @r33] and subject them to a principal components analysis (using ArcMap 10.3.1). We used the first `r as.character(as.english(cs1.params.LST[[MODE]]$n.pc.layers))` principle components (cumulatively explained `r round(cs1.pca.DF[[4]][cs1.params.LST[[MODE]]$n.pc.layers],1)` % of the total variation) to characterise the environmental variation across the study area (Figure 3).

We used four bird species in this case study: blue-winged kookaburra (_Dacelo leachii_), brown-backed honeyeater (_Ramsayornis modestus_), the brown falcon (_Falco berigora_), and the pale headed rosella (_Platycercus adscitus_). We used these bird species because they have different evolutionary histories and life-history strategies. We then mapped the extent of occurrence for each species using the following method: we obtained observation data from the Atlas of Living Australia across the whole of Australia [using the \texttt{ALA4R R} package; @r453], spatially thinned the data to omit points within `r cs1.params.LST[[MODE]]$thin.distance/1000` km$^2$ of each other to ameliorate the effects of sampling bias [using the developmental version of the \texttt{spThin R} package; \url{www.github.com/mlammens/spThin}; @r454], and fit `r cs1.params.LST[[MODE]]$mcp.percent` % minimum convex polygons [using the \texttt{adehabitatHR R} package; @r455]. We used this method because it is reproducible using freely available data and moderately accurate.

We generated `r cs1.params.LST[[MODE]]$dp.number` demand points for each species (Figure S2). To achieve this, for each species, we generated random points inside the species range and at each point extracted the principal component values at that location. We then fit hyperbox kernels to the distribution of these points to characterise the realised niche of each species [using a manually chosen bandwidth of `r cs1.params.LST[[MODE]]$dp.bandwidth[1]` and a `r cs1.params.LST[[MODE]]$dp.quantile` quantile to map the core parts of the species' niches; implemented in the \texttt{hypervolume R} package; @r231]. We then generated uniformly distributed points inside the species' kernels and estimated the density of the training points at the uniformly generated points. These uniformly distributed points and associated density estimated were used as demand point coordinates and weights (respectively).

We generated two prioritisations. The first prioritisation was generated using `r cs1.params.LST[[MODE]]$amount.target*100` % amount-based targets for all species. The second prioritisation was generated the same amount-based target with an additional `r cs1.params.LST[[MODE]]$space.target*100` % niche-based for each species.

### Results

## Case-study 2
### Methods
Here we used of space-based targets to generate a prioritisation to secure a representative sample of a species' intra-specific genetic variation. We used species occurrence and genomic data collected by the international IntraBioDiv project in the European Alps [@r451; see @r452 and @r478 for further explanation of data collection methods]. Although this dataset contains multiple plant species, we used data for the `r  cs2.params.LST[[MODE]]$common.name` (\textit{`r gsub('_', ' ',cs2.params.LST[[MODE]]$species.name,fixed=TRUE)`}). This species was chosen because it provides is locally wide-spread and has been shown to exhibit significant genetic structure [@r454]. Members of the IntraBioDiv project collected data using a $20^{\prime}$ longitude by $21^{\prime}$ latitude grid (approx. 22.3 km $\times$ 25 km; Figure 5a). They visited every second grid cell, and if the `r cs2.params.LST[[MODE]]$common.name` was detected in a cell, samples were collected from three individuals. Samples were genotyped using amplified fragment length polymorphisms [AFLP; @r453], and used to construct matrices denoting the presence/absence of polymorphisms at loci. In total, `r structurer::n.samples(cs2.spp.StructureData.LST[[1]])` individuals were genotyped at `r structurer::n.loci(cs2.spp.StructureData.LST[[1]])` markers.

We used non-metric multi-dimensional scaling [NMDS; using Gower distances to accommodate sparsity; @r459; implemented the \texttt{cluster R} package; @r457] to ordinate the individual-level presence/absence genetic data into two continuous variables [implemented in the \texttt{vegan R} package; @r458]. These continuous variables described the main theme of genetic variation within the species. We calculated the average of the values associated with individuals in each grid cell. These values were used to create a genetic attribute space (Figures 5c and 5d).

The grid cells were used as planning units. The grid-cell averaged ordinations were used to describe the typical genetic characteristics of individuals in the planning unit. Since the number of planning units was relatively small, we used the same grid-cell averaged ordinations as demand points. To ensure that the solutions did not prioritise particularly costly areas that could not be used for conservation action, we included opportunity cost data (Figure 5b). We obtained population density data [obtained at 1 km resolution from the Global Rural-Urban Mapping Project; \texttt{GRUMP V1}; @r456] and estimated the total population density inside each grid cell.

We generated two prioritisations. The first prioritisation was generated using an `r cs2.params.LST[[MODE]]$amount.target*100` % amount-based target. The second prioritisation was generated using the same amount-based target and an additional `r cs2.params.LST[[MODE]]$genetic.target*100` % genetic-based target.

### Results

## Implications and future directions
The \texttt{rapr} R package provides a unified approach to reserve selection. One of the key advantages of the \texttt{rapr R} package is that it is general enough that any spatial variation could be considered an attribute space, regardless of whether this variation is intrinsic or extrinsic to the feature(s). This R package provides decision makers with the tools to generate prioritisations that secure both biodiversity patterns and processes. Additionally, the package contains functionality to accommodate uncertainty in the distribution of features, and also identify suitably connected reserves. Both the simulated and case-study species suggest that conservation planning exercises need to explicitly consider biodiversity processes during the reserve selection process to capture them.

@@

To maximise the long-term persistence of biodiversity--the stated goal of conservation--decision makers need to identify prioritisations that preserve existing patterns of biodiversity and the processes that support them. To achieve this, conservation planners need a decision support tool that can explicitly accommodate biodiversity patterns and processes. Here, we developed the \texttt{rapr R} package to fill this void. By exploring the functionality of this package using several simulated species, we found that including space-based targets can radically change a prioritisation for the simplest of species.
